---
title: "Bayesian network analysis of TCR:pMHC contacts"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load data and filter it

```{r}
library(dplyr)
library(RColorBrewer)
library(gplots)
library(reshape2)
library(ggplot2)
library(bnlearn)
library(pROC)
library(stringr)

rf = colorRampPalette(rev(brewer.pal(11, 'Spectral')))
r = rf(32)

# Energy computation https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2998601/
CDR_BINS = 5
CDR_BIN_SIZE = 1
CDR_ABS_MAXBIN = 5
AG_BINS = 5

get_rel_pos = function(pos, len, bins) {
  # move to middle
  pos = ifelse(len < bins, pos - len/2 + bins/2, pos)
  len = ifelse(len < bins, bins, len)
  
  as.integer(((pos / len) * # 0..(1
    bins) + 1) # 1..bins
}

get_abs_pos = function(pos, len, bin_size) {
  x = as.integer(abs(pos - len / 2) / bin_size + 1)
  ifelse(x > CDR_ABS_MAXBIN, CDR_ABS_MAXBIN, x)
}

df = read.table("structure.txt", header = T, sep="\t") %>%
  filter(tcr_region %in% c("CDR1", "CDR2", "CDR3")) %>%
  mutate(tcr_chain = as.factor(substr(as.character(tcr_v_allele), 1, 3)),
         pos_tcr = as.numeric(pos_tcr),
         len_tcr = as.numeric(len_tcr),
         pos_antigen = as.numeric(pos_antigen),
         len_antigen = as.numeric(len_antigen)) %>%
  dplyr::select(pdb_id, antigen_seq, tcr_region, tcr_chain, 
                pos_tcr, len_tcr, aa_tcr, 
                pos_antigen, len_antigen, aa_antigen, mhc_type, 
                distance,
                energy) %>%
  mutate(contact = as.factor(energy < 0), energy = ifelse(energy > 0, 0, energy),
         aa_pair = as.factor(ifelse(as.character(aa_tcr) < as.character(aa_antigen), paste(aa_tcr, aa_antigen), paste(aa_antigen, aa_tcr))),
         pos_rel_tcr = as.factor(get_rel_pos(pos_tcr, len_tcr, CDR_BINS)),
         pos_abs_tcr = as.factor(get_abs_pos(pos_tcr, len_tcr, CDR_BIN_SIZE)),
         pos_rel_antigen = as.factor(get_rel_pos(pos_antigen, len_antigen, AG_BINS)))

MIN_CONTACTS = 5

good_pdbs = (df %>%
  group_by(pdb_id) %>%
  summarize(contacts = sum(contact == "TRUE")) %>%
  filter(contacts >= MIN_CONTACTS))$pdb_id

df = subset(df, pdb_id %in% good_pdbs)

pdb_id = df$pdb_id
```

```{r}
df.e = df %>% filter(energy < 0) %>%
  group_by(aa_pair) %>%
  summarize(E = mean(energy))

df.e$aa_tcr = str_split_fixed(df.e$aa_pair, " ", 2)[, 1]
df.e$aa_antigen = str_split_fixed(df.e$aa_pair, " ", 2)[ ,2]

df.e2 = df.e
df.e2$aa_tcr = df.e$aa_antigen
df.e2$aa_antigen = df.e$aa_tcr

df.e = rbind(df.e, df.e2)
df.e2 = NULL

ggplot(df.e, aes(x=aa_tcr, y = aa_antigen, fill=E)) +
  geom_tile() + 
  scale_fill_gradientn(colors=colorRampPalette(rev(brewer.pal(11, 'YlOrRd')))(32)) +
  theme_bw()
```

Theoretical model

```{r}
emp_net = model2network(paste(
  "[contact]",
  "[tcr_chain|contact]",
  "[mhc_type|contact]",
  "[tcr_region|contact]",
  "[pos_rel_tcr|contact]", 
  "[pos_abs_tcr|contact]", 
  "[pos_rel_antigen|pos_rel_tcr:tcr_region:tcr_chain:mhc_type:contact]", 
  "[aa_pair|contact]", 
  sep =""))

graphviz.plot(emp_net)
```

Fit conditional probabilities

```{r}
df.train = df %>% dplyr::select(tcr_region, tcr_chain,
                          pos_rel_tcr, pos_abs_tcr, pos_rel_antigen, aa_pair, 
                          mhc_type, 
                          contact) %>%
  droplevels()

head(df.train)

summary(df.train)

fit = bn.fit(emp_net, df.train, method="bayes")

BIC(fit, df.train)
```

Compute the probabilities and plot the ROC curve

```{r}
res = predict(fit, node="contact", method="bayes-lw", 
              data=df.train, prob=T)

p = attributes(res)$prob

plot.roc(df.train[,"contact"], p[2,], ci=T)
```

Here we build a probability matrix from the **bnlearn** results. The results are, for each variable $X_{i}$ the probability $P(X_{i} | \pi_{i})$, where $\pi_{i}$ is the set of parents according to the defined bayes network. We therefore compute the $P(X) = \prod_{i} P(X_{i} | \pi_{i})$

```{r}
get_prob = function(var_name) {
  tmp = as.data.frame(fit[[var_name]]$prob)
  colnames(tmp) = gsub("Var1", "contact", colnames(tmp))
  colnames(tmp) = gsub("Freq", paste("Freq", var_name, sep="."), colnames(tmp))
  tmp
}

prob.matrix = get_prob("contact")

for (var in colnames(df.train)[!(colnames(df.train) %in% c("contact", "pdb_id"))]) {
  prob.matrix = merge(prob.matrix, get_prob(var))
}

prob.matrix$contact = as.logical(prob.matrix$contact)

prob.matrix$P = apply(prob.matrix[,which(grepl("Freq", colnames(prob.matrix)))], 1,
                function(x) prod(x))

prob.matrix = prob.matrix %>% 
  dplyr::select(contact, mhc_type, 
                tcr_chain, tcr_region, aa_pair, pos_rel_tcr, pos_abs_tcr, pos_rel_antigen, P)
```

### Example

Compute contact probabilities and merge them with initial dataset

```{r}
prob.matrix.energy = prob.matrix %>%
  group_by(mhc_type, 
           tcr_chain, tcr_region, aa_pair, pos_rel_tcr, pos_abs_tcr, pos_rel_antigen) %>%
  summarize(P = P[which(contact)] / sum(P))

prob.matrix.energy = merge(prob.matrix.energy, df.e) %>%
  mutate(E = P * E)

df.m = merge(df, prob.matrix.energy)
```

```{r}
df.m.sum = df.m %>%
  group_by(pdb_id, mhc_type, tcr_chain) %>%
  summarize(energy = sum(energy), E = sum(E), contact = sum(contact == "TRUE"), P = sum(P))

ggplot(df.m.sum, aes(-energy, -E)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(mhc_type ~ tcr_chain)

ggplot(df.m.sum, aes(contact, P)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(mhc_type ~ tcr_chain)
```

Write contact probability matrix so it can be used by future generations

```{r}
prob.matrix.energy.out = prob.matrix.energy

prob.matrix.energy.out$pos_rel_tcr = as.integer(prob.matrix.energy.out$pos_rel_tcr)
prob.matrix.energy.out$pos_abs_tcr = as.integer(prob.matrix.energy.out$pos_abs_tcr)
prob.matrix.energy.out$pos_rel_antigen = as.integer(prob.matrix.energy.out$pos_rel_antigen)

write.table(prob.matrix.energy.out, "contact.energies.txt", sep = "\t", quote = F, row.names = F)
```

Contact probabilities from Bayesian model overlayed by positions of known contacts

```{r}
df.m.1 = df.m %>%
  filter(antigen_seq == "GLCTLVAML") %>%
  droplevels()

ggplot(df.m.1, aes(x=pos_tcr, y=pos_antigen)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = E), cex=1.3) +
  geom_point(aes(color=contact)) +
  scale_x_continuous(breaks=0:20) +
  scale_y_continuous(breaks=0:20) +
  scale_fill_gradient("P", 
                      low="#045a8d", high="white") +
  scale_color_manual(values = c(NA, "red")) +
  facet_grid(tcr_chain ~ tcr_region, scales="free", space="free") +
  theme_bw()
```



