---
title: "Bayesian network analysis of TCR:pMHC contacts"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load data and filter it

```{r}
library(dplyr)
library(RColorBrewer)
library(gplots)
library(reshape2)
library(ggplot2)
library(bnlearn)
library(pROC)
rf = colorRampPalette(rev(brewer.pal(11, 'Spectral')))
r = rf(32)

# Energy computation https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2998601/
CDR_BINS = 5
CDR_BIN_SIZE = 1
CDR_ABS_MAXBIN = 5
AG_BINS = 5

get_rel_pos = function(pos, len, bins) {
  # move to middle
  pos = ifelse(len < bins, pos - len/2 + bins/2, pos)
  len = ifelse(len < bins, bins, len)
  
  as.integer(((pos / len) * # 0..(1
    bins) + 1) # 1..bins
}

get_abs_pos = function(pos, len, bin_size) {
  x = as.integer(abs(pos - len / 2) / bin_size + 1)
  ifelse(x > CDR_ABS_MAXBIN, CDR_ABS_MAXBIN, x)
}

df = read.table("structure.txt", header = T, sep="\t") %>%
  filter(tcr_region %in% c("CDR1", "CDR2", "CDR3")) %>%
  mutate(tcr_chain = as.factor(substr(as.character(tcr_v_allele), 1, 3)),
         pos_tcr = as.numeric(pos_tcr),
         len_tcr = as.numeric(len_tcr),
         pos_antigen = as.numeric(pos_antigen),
         len_antigen = as.numeric(len_antigen)) %>%
  dplyr::select(pdb_id, antigen_seq, tcr_region, tcr_chain, 
                pos_tcr, len_tcr, aa_tcr, 
                pos_antigen, len_antigen, aa_antigen, mhc_type, 
                distance,
                energy) %>%
  mutate(contact = as.factor(energy < 0),
         pos_rel_tcr = as.factor(get_rel_pos(pos_tcr, len_tcr, CDR_BINS)),
         pos_abs_tcr = as.factor(get_abs_pos(pos_tcr, len_tcr, CDR_BIN_SIZE)),
         pos_rel_antigen = as.factor(get_rel_pos(pos_antigen, len_antigen, AG_BINS)))

MIN_CONTACTS = 5

good_pdbs = (df %>%
  group_by(pdb_id) %>%
  summarize(contacts = sum(contact == "TRUE")) %>%
  filter(contacts >= MIN_CONTACTS))$pdb_id

df = subset(df, pdb_id %in% good_pdbs)

print(length(unique(df$pdb_id)))

pdb_id = df$pdb_id
df$contact[is.na(df$contact)] = "FALSE"

df.abspos = df

df = df %>% dplyr::select(tcr_region, tcr_chain,
                          pos_rel_tcr, pos_abs_tcr, aa_tcr, pos_rel_antigen, aa_antigen, 
                          mhc_type, 
                          contact) %>%
  droplevels()

head(df)

summary(df)
```

Theoretical model

```{r}
emp_net = model2network(paste(
  "[contact]",
  "[tcr_chain|contact]",
  "[mhc_type|contact]",
  "[tcr_region|contact]",
  "[pos_rel_tcr|contact]", 
  "[pos_abs_tcr|contact]", 
  "[aa_tcr|contact]", 
  "[pos_rel_antigen|pos_rel_tcr:tcr_region:tcr_chain:mhc_type:contact]", 
  "[aa_antigen|aa_tcr:contact]", 
  sep =""))

graphviz.plot(emp_net)
```

Fit conditional probabilities

```{r}
fit = bn.fit(emp_net,
             df, method="bayes")

BIC(fit, df)
```

Compute the probabilities and plot the ROC curve

```{r}
res = predict(fit, node="contact", method="bayes-lw", 
              data=df, prob=T)

p = attributes(res)$prob

rocobj = plot.roc(df[,"contact"], p[2,], ci=T)

rocobj
```

```{r}
rocobj$thresholds[findInterval(0.9, rocobj$specificities)]
```

Estimated vs predicted number of contacts

```{r}
# http://stats.stackexchange.com/questions/177199/success-of-bernoulli-trials-with-different-probabilities
# https://en.wikipedia.org/wiki/Poisson_binomial_distribution

df.cplx = data.frame(pdb_id = pdb_id)
df.cplx$tcr_chain = df$tcr_chain
df.cplx$tcr_region = df$tcr_region
df.cplx$mhc_type = df$mhc_type
df.cplx$contact = as.logical(df$contact)
df.cplx$p = p[2,]

df.cplx.1 = df.cplx %>%
   group_by(pdb_id, tcr_chain, mhc_type) %>%
   summarise(contacts = sum(as.logical(contact)),
             contacts.pred = sum(p),
             contacts.pred.sd = sqrt(sum(p * (1 - p))),
             k = n())
 
ggplot(df.cplx.1, aes(contacts, contacts.pred)) + 
   geom_density2d() +
   geom_abline(slope = 1, intercept = 0, linetype="dashed") +
   geom_point(shape=21) +
   geom_errorbar(aes(ymin = contacts.pred - contacts.pred.sd, ymax = contacts.pred + contacts.pred.sd)) +
   #scale_color_brewer(palette = "Set1") +
   facet_grid(tcr_chain~mhc_type, scales="free", space="free") +
   theme_bw()

# All CDRs linear model

summary(lm(contacts ~ contacts.pred:mhc_type:tcr_chain, df.cplx.1))

df.cplx.2 = df.cplx %>%
   filter(tcr_region == "CDR3") %>%
   group_by(pdb_id, tcr_chain, mhc_type) %>%
   summarise(contacts = sum(as.logical(contact)),
             contacts.pred = sum(p),
             contacts.pred.sd = sqrt(sum(p * (1 - p))))

ggplot(df.cplx.2, aes(contacts, contacts.pred)) + 
   geom_density2d() +
   geom_abline(slope = 1, intercept = 0, linetype="dashed") +
   geom_point(shape=21) +
   geom_errorbar(aes(ymin = contacts.pred - contacts.pred.sd, ymax = contacts.pred + contacts.pred.sd)) +
   #scale_color_brewer(palette = "Set1") +
   facet_grid(tcr_chain~mhc_type, scales="free", space="free") +
   theme_bw()

# CDR3 only linear model

summary(lm(contacts ~ contacts.pred:mhc_type:tcr_chain, df.cplx.2))
```

Here we build a probability matrix from the **bnlearn** results. The results are, for each variable $X_{i}$ the probability $P(X_{i} | \pi_{i})$, where $\pi_{i}$ is the set of parents according to the defined bayes network. We therefore compute the $P(X) = \prod_{i} P(X_{i} | \pi_{i})$

```{r}
get_prob = function(var_name) {
  tmp = as.data.frame(fit[[var_name]]$prob)
  colnames(tmp) = gsub("Var1", "contact", colnames(tmp))
  colnames(tmp) = gsub("Freq", paste("Freq", var_name, sep="."), colnames(tmp))
  tmp
}

prob.matrix = get_prob("contact")

for (var in colnames(df)[!(colnames(df) %in% c("contact", "pdb_id"))]) {
  prob.matrix = merge(prob.matrix, get_prob(var))
}

prob.matrix$contact = as.logical(prob.matrix$contact)

prob.matrix$P = apply(prob.matrix[,which(grepl("Freq",colnames(prob.matrix)))], 1,
                function(x) prod(x))

prob.matrix = prob.matrix %>% 
  dplyr::select(contact, mhc_type, 
                tcr_chain, tcr_region, aa_tcr, aa_antigen, pos_rel_tcr, pos_abs_tcr, pos_rel_antigen, P)
```

### Example

Compute contact probabilities and merge them with initial dataset

```{r}
prob.matrix.cont = prob.matrix %>%
  group_by(mhc_type, 
           tcr_chain, tcr_region, aa_tcr, aa_antigen, pos_rel_tcr, pos_abs_tcr, pos_rel_antigen) %>%
  summarize(P = P[which(contact)] / sum(P))

df.m = merge(df.abspos, prob.matrix.cont)
```

Write contact probability matrix so it can be used by future generations

```{r}
prob.matrix.cont.out = prob.matrix.cont

prob.matrix.cont.out$pos_rel_tcr = as.integer(prob.matrix.cont.out$pos_rel_tcr)
prob.matrix.cont.out$pos_abs_tcr = as.integer(prob.matrix.cont.out$pos_abs_tcr)
prob.matrix.cont.out$pos_rel_antigen = as.integer(prob.matrix.cont.out$pos_rel_antigen)

write.table(prob.matrix.cont.out, "contact.probs.txt", sep = "\t", quote = F, row.names = F)
```

Contact probabilities from Bayesian model overlayed by positions of known contacts

```{r}
df.m.1 = df.m %>%
  filter(antigen_seq == "GLCTLVAML") %>%
  droplevels()

ggplot(df.m.1, aes(x=pos_tcr, y=pos_antigen)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = P), cex=1.3) +
  geom_point(aes(color=contact)) +
  scale_x_continuous(breaks=0:20) +
  scale_y_continuous(breaks=0:20) +
  scale_fill_gradient("P", #limits = c(0, 1),
                                 low="white",high="#313695") +
  scale_color_manual(values = c(NA, "red")) +
  facet_grid(tcr_chain ~ tcr_region, scales="free", space="free") +
  theme_bw()
```

### Amino acid - amino acid contact preferences

```{r}
prob.aTaAC = prob.matrix %>%
  #filter(tcr_region == "CDR3") %>%
  #mutate(P = P / sum(P)) %>%
  group_by(aa_tcr, aa_antigen, contact) %>%
  summarise(P = sum(P)) %>%
  group_by(aa_tcr, aa_antigen) %>%
  summarise(P = P[which(contact)] / sum(P))
```

```{r}
aa_pair_mat = dcast(prob.aTaAC, aa_tcr ~ aa_antigen)
rownames(aa_pair_mat) = aa_pair_mat$aa_tcr
aa_pair_mat$aa_tcr = NULL
aa_pair_mat = as.matrix(aa_pair_mat)
aa_pair_mat[is.na(aa_pair_mat)] = 0

df.hydro <- data.frame(
  aa = strsplit("I V L F C M A W G T S Y P H N D Q E K R", " ")[[1]],
  hydrop = strsplit("4.5 4.2 3.8 2.8 2.5 1.9 1.8 -0.9 -0.4 -0.7 -0.8 -1.3 -1.6 -3.2 -3.5 -3.5 -3.5 -3.5 -3.9 -4.5", " ")[[1]]
)

df.hydro = df.hydro %>%
  mutate(hydrop = as.numeric(as.character(hydrop))) %>%
  arrange(hydrop) %>%
  mutate(hydrop.sc = round(100 * (hydrop - min(hydrop)) / (max(hydrop) - min(hydrop))))

df.hydro$color = colorRampPalette(brewer.pal(11, 'PRGn'))(101)[df.hydro$hydrop.sc + 1]

aa_colors = df.hydro$color
names(aa_colors) = df.hydro$aa

js_calc = function(p, q) {
  m <- 0.5 * (p + q)
  0.5 * (sum(p * log(p / m)) + sum(q * log(q / m)))
}

js_dist = function(x) {
    mat <- x
    for(i in 1:nrow(mat)) {
        for(j in 1:nrow(mat)) {
            mat[i, j] <- js_calc(x[i, ], x[j, ])
    }}
    return(as.dist(mat))
}

heatmap.2(aa_pair_mat, 
          #hclustfun = function(x) hclust(x, method = "ward"),
          distfun = function(x) js_dist(x),
          RowSideColors = aa_colors[rownames(aa_pair_mat)], 
          ColSideColors = aa_colors[colnames(aa_pair_mat)],
          trace = "none",
          #breaks = seq(0, 0.2, length.out = 101),
          col=colorRampPalette(rev(brewer.pal(11, 'RdBu')))(100))
```

Symmetric matrix

```{r}
#aa_pair_mat_s = aa_pair_mat
#if (!("C" %in% rownames(aa_pair_mat))) {
#  aa_pair_mat_s = rbind(aa_pair_mat_s, matrix(nrow = 1, ncol = ncol(aa_pair_mat), data=0))
#}
aa_pair_mat_s = 0.5 * (aa_pair_mat + t(aa_pair_mat))

heatmap.2(aa_pair_mat_s, 
          #hclustfun = function(x) hclust(x, method = "ward"),
          distfun = function(x) js_dist(x),
          RowSideColors = aa_colors, 
          ColSideColors = aa_colors,
          trace = "none",
          #breaks = seq(0, 0.2, length.out = 101),
          col=colorRampPalette(rev(brewer.pal(11, 'RdBu')))(100))
```

Probability of contact

```{r}
df.1 = prob.matrix %>%
  #filter(tcr_region == "CDR3") %>%
  #mutate(P = P / sum(P)) %>%
  group_by(aa_tcr, contact) %>%
  summarise(P = sum(P)) %>%
  group_by(aa_tcr) %>%
  summarise(P = P[which(contact)] / sum(P)) 

df.2 = prob.matrix %>%
  group_by(aa_antigen, contact) %>%
  summarise(P = sum(P)) %>%
  group_by(aa_antigen) %>%
  summarise(P = P[which(contact)] / sum(P)) 

colnames(df.1) = c("aa", "P_TCR")
colnames(df.2) = c("aa", "P_AG")

df.1 = merge(df.1, df.2)
df.1 = merge(df.1, df.hydro)

ggplot(df.1, aes(x=P_TCR, y=P_AG, fill=hydrop)) +
  geom_label(aes(label=aa)) +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(9, 'PRGn'))(20)) +
  #scale_color_gradientn(colors = df.1$color) + 
  theme_bw()
```

### Positional contact preferences

Contact probability for relative CDR-antigen residue positions

```{r}
df.1 = prob.matrix %>%
  group_by(mhc_type,
           pos_rel_antigen, pos_rel_tcr, tcr_chain, tcr_region, contact) %>%
  summarise(P = sum(P)) %>%
  group_by(mhc_type,
           pos_rel_antigen, pos_rel_tcr, tcr_chain, tcr_region) %>%
  summarise(P = P[which(contact)] / sum(P)) 

ggplot(df.1, aes(x = pos_rel_antigen, y = pos_rel_tcr, fill=P)) +
  geom_tile() + 
  scale_fill_gradientn(colors=r) +
  facet_grid(tcr_region~mhc_type+tcr_chain)
```

```{r}
df.2 = prob.matrix %>%
  group_by(pos_abs_tcr, tcr_region, contact) %>%
  summarise(P = sum(P)) %>%
  group_by(pos_abs_tcr, tcr_region) %>%
  summarise(P = P[which(contact)] / sum(P)) 

ggplot(df.2, aes(x = pos_abs_tcr, y = P, group = tcr_region, color=tcr_region)) +
  geom_point() + geom_line()
```

CDR amino acid + position contact probability

```{r}
df.1 = prob.matrix %>%
  group_by(aa_tcr, pos_rel_tcr, tcr_chain, tcr_region, contact) %>%
  summarise(P = sum(P)) %>%
  group_by(aa_tcr, pos_rel_tcr, tcr_chain, tcr_region) %>%
  summarise(P = P[which(contact)] / sum(P)) 

ggplot(subset(df.1, aa_tcr != "C"), aes(x=pos_rel_tcr, y = aa_tcr, fill=P)) +
  geom_tile() + 
  scale_fill_gradientn(colors=r) +
  facet_grid(tcr_chain ~ tcr_region)
```

Contact energies

```{r}
df.e = df.abspos %>%
  filter(energy < 0) %>%
  select(aa_tcr, aa_antigen, energy) 

df.e2 = df.e
df.e2$aa_tcr = df.e2$aa_antigen
df.e2$aa_antigen = df.e$aa_tcr

df.e = rbind(df.e, df.e2) %>%
  group_by(aa_tcr, aa_antigen) %>%
  summarize(E = mean(energy))

ggplot(df.e, aes(x=aa_tcr, y = aa_antigen, fill=E)) +
  geom_tile() + 
  scale_fill_gradientn(colors=colorRampPalette(rev(brewer.pal(11, 'YlOrRd')))(32))

aa_pair_mat1 = dcast(df.e, aa_tcr ~ aa_antigen)
rownames(aa_pair_mat1) = aa_pair_mat1$aa_tcr
aa_pair_mat1$aa_tcr = NULL
aa_pair_mat1 = as.matrix(aa_pair_mat1)
aa_pair_mat1[is.na(aa_pair_mat1)] = 0

heatmap.2(aa_pair_mat1)

write.table(df.e, "energy.mean.txt", sep = "\t", quote = F, row.names = F)
```

```{r}
df.2 = merge(prob.matrix, df.e) %>%
  group_by(aa_tcr, aa_antigen, pos_rel_tcr, pos_abs_tcr, tcr_chain, tcr_region, E, contact) %>%
  summarise(P = sum(P)) %>%
  group_by(aa_tcr, aa_antigen, pos_rel_tcr, pos_abs_tcr, tcr_chain, tcr_region, E) %>%
  summarise(P = P[which(contact)] / sum(P))  %>%
  group_by(aa_tcr, pos_rel_tcr, pos_abs_tcr, tcr_chain, tcr_region) %>%
  summarise(energy = mean(P * E)) # all antigen aa probs are equal

df.2$pos_rel_tcr = as.integer(df.2$pos_rel_tcr)
df.2$pos_abs_tcr = as.integer(df.2$pos_abs_tcr)

write.table(df.2, "energy.tcr.txt", sep = "\t", quote = F, row.names = F)
```
