---
title: "TCR-only contact model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(RColorBrewer)
library(gplots)
library(reshape2)
library(ggplot2)
library(bnlearn)
library(pROC)

DIST_THRESHOLD = 6
ENERGY_THRESHOLD = 0
CDR_BINS = 5

df = read.table("structure.txt", header = T, sep="\t") %>%
  filter(tcr_region %in% c("CDR1", "CDR2", "CDR3")) %>%
  mutate(tcr_chain = as.factor(substr(as.character(tcr_v_allele), 1, 3)),
         pos_tcr = as.numeric(pos_tcr),
         len_tcr = as.numeric(len_tcr),
         pos_antigen = as.numeric(pos_antigen),
         len_antigen = as.numeric(len_antigen)) %>%
  dplyr::select(pdb_id, antigen_seq, tcr_region, tcr_chain, 
                pos_tcr, len_tcr, aa_tcr, 
                mhc_type, energy) %>%
  mutate(contact = as.factor(energy < ENERGY_THRESHOLD),
         pos_rel_tcr = cut(pos_tcr / (len_tcr - 1), CDR_BINS)) #%>%
  #group_by(pdb_id, antigen_seq, tcr_region, tcr_chain, 
  #              pos_tcr, len_tcr, aa_tcr, 
  #              mhc_type, pos_rel_tcr) %>%
  #summarize(contact = as.factor(sum(contact) > 0))

print(length(unique(df$pdb_id)))

pdb_id = df$pdb_id

df.abspos = df

df = as.data.frame(df) %>% dplyr::select(tcr_region, tcr_chain,
                          pos_rel_tcr, aa_tcr, 
                          mhc_type, contact) %>%
  droplevels()

df$contact[is.na(df$contact)] = "FALSE"

head(df)

summary(df)
```

Inferred model

```{r}
tb = tree.bayes(df, training = "contact")

graphviz.plot(tb)
```

Theoretical model

```{r}
emp_net = model2network(paste(
  "[contact]",
  "[tcr_chain|contact]",
  "[mhc_type|contact]",
  "[tcr_region|contact]",
  "[pos_rel_tcr|tcr_region:tcr_chain:mhc_type:contact]", 
  "[aa_tcr|contact]",
  sep =""))

graphviz.plot(emp_net)
```

Fit conditional probabilities

```{r}
fit = bn.fit(emp_net,
             df, method="bayes")

BIC(fit, df)
```

Compute the probabilities and plot the ROC curve

```{r}
res = predict(fit, node="contact", method="bayes-lw", 
              data=df, prob=T)

p = attributes(res)$prob

rocobj = plot.roc(df[,"contact"], p[2,], ci=T)

rocobj
```

Estimated vs predicted number of contacts

```{r}
df.cplx = data.frame(pdb_id = pdb_id)
df.cplx$tcr_chain = df$tcr_chain
df.cplx$tcr_region = df$tcr_region
df.cplx$mhc_type = df$mhc_type
df.cplx$contact = as.logical(df$contact)
df.cplx$p = p[2,]

df.cplx.1 = df.cplx %>%
   group_by(pdb_id, tcr_chain, mhc_type) %>%
   summarise(contacts = sum(as.logical(contact)),
             contacts.pred = sum(p))
 
ggplot(df.cplx.1, aes(contacts, contacts.pred, color = tcr_chain)) + 
   geom_abline(slope = 1, intercept = 0, linetype="dashed") +
   geom_point(shape=21) +
   scale_x_continuous(limits=c(0, 20)) +
   scale_y_continuous(limits=c(0, 20)) +
   scale_color_brewer(palette = "Set1") +
   facet_wrap(~mhc_type) +
   theme_bw()

# All CDRs linear model

summary(lm(contacts ~ contacts.pred, df.cplx.1))

df.cplx.2 = df.cplx %>%
   filter(tcr_region == "CDR3") %>%
   group_by(pdb_id, tcr_chain, mhc_type) %>%
   summarise(contacts = sum(as.logical(contact)),
             contacts.pred = sum(p))

# CDR3 only linear model

summary(lm(contacts ~ contacts.pred, df.cplx.2))
```

Here we build a probability matrix from the **bnlearn** results. The results are, for each variable $X_{i}$ the probability $P(X_{i} | \pi_{i})$, where $\pi_{i}$ is the set of parents according to the defined bayes network. We therefore compute the $P(X) = \prod_{i} P(X_{i} | \pi_{i})$

```{r}
get_prob = function(var_name) {
  tmp = as.data.frame(fit[[var_name]]$prob)
  colnames(tmp) = gsub("Var1", "contact", colnames(tmp))
  colnames(tmp) = gsub("Freq", paste("Freq", var_name, sep="."), colnames(tmp))
  tmp
}

prob.matrix = get_prob("contact")

for (var in colnames(df)[!(colnames(df) %in% c("contact", "pdb_id"))]) {
  prob.matrix = merge(prob.matrix, get_prob(var))
}

prob.matrix$contact = as.logical(prob.matrix$contact)

prob.matrix$P = apply(prob.matrix[,which(grepl("Freq",colnames(prob.matrix)))], 1,
                function(x) prod(x))

prob.matrix = prob.matrix %>% 
  dplyr::select(contact, mhc_type, tcr_chain, tcr_region, aa_tcr, pos_rel_tcr, P)

prob.matrix.cont = prob.matrix %>%
  group_by(mhc_type, tcr_chain, tcr_region, aa_tcr, pos_rel_tcr) %>%
  summarize(P = P[which(contact)] / sum(P))
```

```{r}
write.table(prob.matrix.cont, "contact.probs.tcr.txt", sep = "\t", quote = F, row.names = F)
```

```{r}
rf = colorRampPalette(rev(brewer.pal(11, 'Spectral')))
r = rf(32)

ggplot(prob.matrix.cont, aes(x=pos_rel_tcr, y = aa_tcr, fill=P)) +
  geom_tile() + 
  scale_fill_gradientn(colors=r) +
  facet_grid(tcr_region ~ tcr_chain + mhc_type)
```

