---
title: "A local GLM-based pairwise contact model for TCR:pMHC complexes"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---

```{r}

library(data.table)
library(ggplot2)
library(reshape2)
library(dplyr)
library(ggbeeswarm)
library(RColorBrewer)
library(pROC)
library(stringr)
library(gplots)
library(gridExtra)
```


```{r}
df = fread("../../result/structure.txt", header=T, sep="\t")[tcr_region %in% c("CDR1", "CDR2", "CDR3")]
  
df$tcr_chain = as.factor(substr(as.character(df$tcr_v_allele), 1, 3))
df$contact = df$distance <= 4.5 
df$aa_pair = with(df, 
  as.factor(ifelse(as.character(aa_tcr) < as.character(aa_antigen), paste(aa_tcr, aa_antigen, sep = "_"), paste(aa_antigen, aa_tcr, sep = "_"))))

summary(df)
```

# Some EDA

## Contact distribution

Contacts by MHC, chain and CDR

```{r}
df.contact.sum = df[,.(contacts = sum(contact)),by=.(pdb_id, tcr_chain, tcr_region, mhc_type)]

df.contact.sum.pdb = df.contact.sum[,.(contacts = sum(contacts)), by=.(pdb_id, mhc_type)][contacts>5]

ggplot(df.contact.sum.pdb, aes(contacts, color = mhc_type)) + 
  geom_freqpoly() + 
  scale_color_brewer(palette = "Set1") +
  theme_bw()

df.contact.sum = df.contact.sum[pdb_id %in% df.contact.sum.pdb$pdb_id]

ggplot(df.contact.sum, aes(x=tcr_region, group = tcr_region, y = contacts, fill = tcr_region)) +
  geom_violin() +
  scale_fill_brewer(guide = F, "TCR region", palette = "Set1") +
  xlab("") + ylab("Antigen contacts (closest atom distance < 4.5Å)") +
  facet_grid(tcr_chain~mhc_type) +
  theme_bw()

a = aov(contacts~tcr_chain*tcr_region*mhc_type, df.contact.sum)
anova(a)

TukeyHSD(a, "tcr_region")
TukeyHSD(a, "tcr_chain")
TukeyHSD(a, "tcr_chain:tcr_region")
```

Filter TCRs with no contacts

```{r}
df = df[pdb_id %in% df.contact.sum.pdb$pdb_id ]
```

Contact distribution on antigen

```{r}
df.contact.dist.ag = df[,
                        .(contacts = sum(contact), total.pdb = length(unique(pdb_id))),
                        by=.(tcr_chain, tcr_region, mhc_type, pos_antigen, len_antigen)]

ggplot(df.contact.dist.ag, aes(x = pos_antigen - len_antigen / 2, weight = contacts / total.pdb, color = tcr_chain)) +
  geom_freqpoly(bins=10) +
  facet_grid(mhc_type~tcr_region) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()

ggplot(df.contact.dist.ag, aes(x = (pos_antigen - len_antigen / 2) / len_antigen, weight = contacts / total.pdb, color = tcr_chain)) +
  geom_freqpoly(bins=10) +
  facet_grid(mhc_type~tcr_region) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()
```

vs antigen length

```{r}
df.contact.dist.ag.len = df[,
                        .(contacts = sum(contact), total.pdb = length(unique(pdb_id))),
                        by=.(pos_antigen, len_antigen, mhc_type)]

ggplot(df.contact.dist.ag.len, aes(x = (pos_antigen - len_antigen / 2), group = paste(len_antigen,mhc_type), 
                                   weight = contacts / total.pdb, color = mhc_type)) +
  geom_freqpoly(bins=10) +
  facet_wrap(~len_antigen) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()
```

Contact distribution on TCR

```{r}
df.contact.dist.tcr = df[,
                        .(contacts = sum(contact), total.pdb = length(unique(pdb_id))),
                        by=.(tcr_chain, tcr_region, mhc_type, pos_tcr, len_tcr)]

ggplot(df.contact.dist.tcr, aes(x = pos_tcr - len_tcr / 2, weight = contacts / total.pdb, color = tcr_chain)) +
  geom_freqpoly(bins=10) +
  facet_grid(mhc_type~tcr_region) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()

ggplot(df.contact.dist.tcr, aes(x = (pos_tcr - len_tcr / 2) / len_tcr, weight = contacts / total.pdb, color = tcr_chain)) +
  geom_freqpoly(bins=10) +
  facet_grid(mhc_type~tcr_region) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()
```

vs CDR3 len

```{r}
df.contact.dist.tcr.len = df[tcr_region == "CDR3",
                        .(contacts = sum(contact), total.pdb = length(unique(pdb_id))),
                        by=.(pos_tcr, len_tcr)]

ggplot(df.contact.dist.tcr.len, aes(x = (pos_tcr - len_tcr / 2), group = len_tcr, weight = contacts / total.pdb)) +
  geom_freqpoly(bins=10) +
  facet_wrap(~len_tcr) +
  theme_bw()
```

> Different chain residues contact the same AG residue, CDR1-3 residues contact MHC, this should go to general EDA

# Modelling

Center CDR and antigen residue indices so that the central residue has 0 index

```{r}
df.pred = df
df.pred$pos_tcr_c = with(df.pred, pos_tcr - round(len_tcr/2))
df.pred$pos_antigen_c = with(df.pred, pos_antigen - round(len_antigen/2))
```

## Calpha distance model

### Simple mean model

Mean Calpha distances for centered coordinates

```{r}
df.ca.mean = df.pred[,.(distance_CA.m = mean(distance_CA)), 
                        by=.(tcr_chain, tcr_region, mhc_type, pos_tcr_c, pos_antigen_c)]

ggplot(df.ca.mean, aes(x=pos_tcr_c, y=pos_antigen_c, fill=distance_CA.m)) +
  geom_tile() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "white") + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "white") + 
  facet_grid(mhc_type~tcr_chain+tcr_region) +
  scale_x_continuous(limits=c(-10,10)) + 
  scale_y_continuous(limits=c(-10,10)) + 
  scale_fill_gradientn(colors=colorRampPalette(brewer.pal(11, 'Spectral'))(32), limits=c(5, 20)) +
  theme_bw() + 
  theme(panel.background = element_rect(fill = 'grey50'),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

> TODO: Some smoothing/bootstrap can go here

### Checking the model

Add mean distance values

```{r}
df.pred = df.pred[as.data.table(df.ca.mean), on = .(tcr_chain, tcr_region, mhc_type, pos_tcr_c, pos_antigen_c)]
```

Compare to true Calpha distance values

```{r}
ggplot(df.pred, aes(x=round(distance_CA.m), group = round(distance_CA.m), y = distance_CA)) +
  geom_boxplot() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(limits=c(5,20)) +
  scale_y_continuous(limits=c(5,20)) +
  facet_grid(mhc_type~tcr_chain+tcr_region) +
  theme_bw()

summary(lm(distance_CA ~ distance_CA.m, df.pred))
```

Plot distance distribution for contacts and non-contacts, for real and estimated distances:

```{r}
ggplot(df.pred, aes(x = distance_CA, fill = contact)) +
  geom_histogram(binwidth = 1) +
  facet_grid(mhc_type~tcr_chain+tcr_region, scales="free_y") +
  scale_x_continuous(limits=c(0,20))+
  scale_fill_brewer(palette = "Set1") +
  theme_bw()

ggplot(df.pred, aes(x = distance_CA.m, fill = contact)) +
  geom_histogram(binwidth = 1) +
  facet_grid(mhc_type~tcr_chain+tcr_region, scales="free_y") +
  scale_x_continuous(limits=c(0,20))+
  scale_fill_brewer(palette = "Set1") +
  theme_bw()
```

## Amino acid preferences and Calpha distance

Using a generalized linear model (GLM) to fit contacts, operate with amino acid pairs, ignoring which one is in the TCR and which comes from antigen. 

```{r}
res = glm(contact ~ distance_CA + aa_pair + 0, family = binomial(), data = df.pred)

summary(res)
```

### Explore contact probability coefficients

Lets look at GLM coefficients for each amino acid pair. The higher the coefficient - the bigger contact probaility given equal Calpha distance

```{r}
coef = res$coefficients
pvalue = coef(summary(res))[,4]
coef = ifelse(pvalue < 0.05, coef, NA)
names(coef) = str_split_fixed(names(coef), "aa_pair", 2)[,2]

df.aa.coef = data.frame(coef = coef,
                     aa_tcr = str_split_fixed(names(coef), "_", 2)[, 1],
                     aa_antigen = str_split_fixed(names(coef), "_", 2)[ ,2]) %>%
  filter(aa_tcr != "" & aa_antigen != "") %>%
  droplevels

df.aa.coef.diag = df.aa.coef
df.aa.coef.diag$aa_pair = with(df.aa.coef.diag, 
  as.factor(ifelse(as.character(aa_tcr) < as.character(aa_antigen), paste(aa_tcr, aa_antigen, sep = "_"), paste(aa_antigen, aa_tcr, sep = "_")))) 

df.aa.coef.diag = df.aa.coef.diag %>% select(aa_pair, coef)

df.aa.coef.rev = df.aa.coef
df.aa.coef.rev$aa_tcr = df.aa.coef$aa_antigen
df.aa.coef.rev$aa_antigen = df.aa.coef$aa_tcr

df.aa.coef = rbind(df.aa.coef, df.aa.coef.rev) %>% unique()

# transform to matrix and plot heatmap.2

aa_pair_mat = dcast(df.aa.coef, aa_tcr ~ aa_antigen, value.var = "coef", fun.aggregate = mean)
rownames(aa_pair_mat) = aa_pair_mat$aa_tcr
aa_pair_mat$aa_tcr = NULL
aa_pair_mat = as.matrix(aa_pair_mat)

df.hydro <- data.frame(
  aa = strsplit("I V L F C M A W G T S Y P H N D Q E K R", " ")[[1]],
  hydrop = strsplit("4.5 4.2 3.8 2.8 2.5 1.9 1.8 -0.9 -0.4 -0.7 -0.8 -1.3 -1.6 -3.2 -3.5 -3.5 -3.5 -3.5 -3.9 -4.5", " ")[[1]]
)

df.hydro = df.hydro %>%
  mutate(hydrop = as.numeric(as.character(hydrop))) %>%
  arrange(hydrop) %>%
  mutate(hydrop.sc = round(100 * (hydrop - min(hydrop)) / (max(hydrop) - min(hydrop))))

df.hydro$color = colorRampPalette(brewer.pal(11, 'PRGn'))(101)[df.hydro$hydrop.sc + 1]

aa_colors = df.hydro$color
names(aa_colors) = df.hydro$aa

heatmap.2(aa_pair_mat, 
          hclustfun = function(x) hclust(x, method = "ward.D2"),
          RowSideColors = aa_colors[rownames(aa_pair_mat)], 
          ColSideColors = aa_colors[colnames(aa_pair_mat)],
          trace = "none",
          #breaks = seq(-16, -7, length.out = 101),
          col=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32))
```

### Contact probability GLM coefficient imputation

Generate data for the imputation and EDA. Here we model coefficients with Kidera factors and check the correlation between Kidera factor values

> TODO: Vadim, perhaps we can allow interaction here. I mean right now we have Kidera_1 + Kidera_2, I would like to see Kidera_1 + Kidera_2 + Kidera_1 * Kidera_2, this makes more sense for values like hydrophobicity or charge.

```{r, fig.width=10}
kidera = t(data.frame(lapply(strsplit("A,-1.56,-1.67,-0.97,-0.27,-0.93,-0.78,-0.2,-0.08,0.21,-0.48;R,0.22,1.27,1.37,1.87,-1.7,0.46,0.92,-0.39,0.23,0.93;N,1.14,-0.07,-0.12,0.81,0.18,0.37,-0.09,1.23,1.1,-1.73;D,0.58,-0.22,-1.58,0.81,-0.92,0.15,-1.52,0.47,0.76,0.7;C,0.12,-0.89,0.45,-1.05,-0.71,2.41,1.52,-0.69,1.13,1.1;Q,-0.47,0.24,0.07,1.1,1.1,0.59,0.84,-0.71,-0.03,-2.33;E,-1.45,0.19,-1.61,1.17,-1.31,0.4,0.04,0.38,-0.35,-0.12;G,1.46,-1.96,-0.23,-0.16,0.1,-0.11,1.32,2.36,-1.66,0.46;H,-0.41,0.52,-0.28,0.28,1.61,1.01,-1.85,0.47,1.13,1.63;I,-0.73,-0.16,1.79,-0.77,-0.54,0.03,-0.83,0.51,0.66,-1.78;L,-1.04,0,-0.24,-1.1,-0.55,-2.05,0.96,-0.76,0.45,0.93;K,-0.34,0.82,-0.23,1.7,1.54,-1.62,1.15,-0.08,-0.48,0.6;M,-1.4,0.18,-0.42,-0.73,2,1.52,0.26,0.11,-1.27,0.27;F,-0.21,0.98,-0.36,-1.43,0.22,-0.81,0.67,1.1,1.71,-0.44;P,2.06,-0.33,-1.15,-0.75,0.88,-0.45,0.3,-2.3,0.74,-0.28;S,0.81,-1.08,0.16,0.42,-0.21,-0.43,-1.89,-1.15,-0.97,-0.23;T,0.26,-0.7,1.21,0.63,-0.1,0.21,0.24,-1.15,-0.56,0.19;W,0.3,2.1,-0.72,-1.57,-1.16,0.57,-0.48,-0.4,-2.3,-0.6;Y,1.38,1.48,0.8,-0.56,0,-0.68,-0.31,1.03,-0.05,0.53;V,-0.74,-0.71,2.04,-0.4,0.5,-0.81,-1.07,0.06,-0.46,0.65", ";")[[1]], function (x) { strsplit(x, ",")[[1]] } ))); kidera = data.frame(kidera[,1], apply(kidera[, 2:11], 2, as.numeric)); kidera[,1] = as.character(kidera[,1]); row.names(kidera) = kidera[,1]; names(kidera) = c("amino.acid", "f1", "f2", "f3", "f4", "f5", "f6", "f7", "f8", "f9", "f10")

#kidera[,2:ncol(kidera)] = lapply(kidera[,2:ncol(kidera)], function (col) (col - min(col)) / (max(col) - min(col)))



generate_data <- function (.mat, .train_size = .7, .cv = 10, .seed = 42) {
  melted = melt(.mat)[melt(upper.tri(.mat, diag = T))[,3],]
  melted[,1] = as.character(melted[,1])
  melted[,2] = as.character(melted[,2])
  train_data = melted[!is.na(melted[,3]), ]
  test_data = melted[is.na(melted[,3]), ]
  
  med = median(train_data[,3])
  
  hi_logic = train_data[,3] >= med
  lo_logic = train_data[,3] < med
  
  train_size = round(.train_size * nrow(train_data))
  
  train_inds = list()
  val_inds = list()
  
  set.seed(.seed)
  for (i in 1:.cv) {
    hi_inds = sample(which(hi_logic), train_size / 2, F)
    lo_inds = sample(which(lo_logic), train_size / 2, F)
    train_inds[[i]] = sample(c(hi_inds, lo_inds))
    
    val_inds[[i]] = sample(c(setdiff(which(hi_logic), hi_inds), setdiff(which(lo_logic), lo_inds)))
  }
  
  res = matrix(0, nrow(train_data), 10)
  for (i in 1:nrow(res)) {
    res[i,] = (kidera[train_data[i,1], ] + kidera[train_data[i,2], ]) / 2
  }
  row.names(res) = paste0(train_data[,1], train_data[,2])
  
  res_tst = matrix(0, nrow(test_data), 10)
  for (i in 1:nrow(res_tst)) {
    res_tst[i,] = (kidera[test_data[i,1], ] + kidera[test_data[i,2], ]) / 2
  }
  row.names(res_tst) = paste0(test_data[,1], test_data[,2])
  
  list(X = res, y = train_data[,3], X_test = res_tst, train = train_inds, val = val_inds)
}


aa_data = generate_data(aa_pair_mat, .cv = 10)

suppressWarnings( {
  ps = lapply(1:10, function (i) { qplot(aa_data$X[,i], aa_data$y, geom = c("point", "smooth"), method="lm") + theme_bw() })
  })
summary(lm(aa_data$y ~ aa_data$X))
do.call(gridExtra::grid.arrange, c(ps, list(ncol = 5)))
```

Impute the missing values in original contact probability coefficient matrix

```{r}
impute_knn_mean <- function (X_train, y_train, X_val, y_val, X_test, .k) {
  res_tr = sapply(1:nrow(X_train), function (row_i) { 
    tmp = sapply(1:nrow(X_train), function (row_j) {
      if (row_i != row_j) { sqrt(sum((X_train[row_i, ] - X_train[row_j, ]) ^ 2))
      } 
      else { 20 }
    })
    mean(y_train[order(tmp)[1:.k]])
  })
  
  res_val = sapply(1:nrow(X_val), function (row_i) { 
    tmp = sapply(1:nrow(X_train), function (row_j) {
      sqrt(sum((X_val[row_i, ] - X_train[row_j, ]) ^ 2))
    })
    mean(y_train[order(tmp)[1:.k]])
  })
  
  merged = rbind(X_train, X_val)
  merged_y = c(y_train, y_val)
  imp_tst = sapply(1:nrow(X_test), function (row_i) { 
    tmp = sapply(1:nrow(merged), function (row_j) {
      sqrt(sum((X_test[row_i, ] - merged[row_j, ]) ^ 2))
    })
    mean(merged_y[order(tmp)[1:.k]])
  })
  
  list(res_tr, res_val, imp_tst)
}


impute_knn_dist <- function (X_train, y_train, X_val, y_val, X_test, .k) {
  res_tr = sapply(1:nrow(X_train), function (row_i) { 
    tmp = sapply(1:nrow(X_train), function (row_j) {
      if (row_i != row_j) { sqrt(sum((X_train[row_i, ] - X_train[row_j, ]) ^ 2)) } 
      else { 20 }
    })
    mean((min(tmp[order(tmp)[1:.k]]) / tmp[order(tmp)[1:.k]]) ^ 2 * y_train[order(tmp)[1:.k]])
  })
  
  res_val = sapply(1:nrow(X_val), function (row_i) { 
    tmp = sapply(1:nrow(X_train), function (row_j) {
      sqrt(sum((X_val[row_i, ] - X_train[row_j, ]) ^ 2))
    })
    mean((min(tmp[order(tmp)[1:.k]]) / tmp[order(tmp)[1:.k]]) ^ 3 * y_train[order(tmp)[1:.k]])
  })
  
  merged = rbind(X_train, X_val)
  merged_y = c(y_train, y_val)
  imp_tst = sapply(1:nrow(X_test), function (row_i) { 
    tmp = sapply(1:nrow(merged), function (row_j) {
      sqrt(sum((X_test[row_i, ] - merged[row_j, ]) ^ 2))
    })
    mean(merged_y[order(tmp)[1:.k]])
  })
  
  list(res_tr, res_val, imp_tst)
}


impute_reg <- function (X_train, y_train, X_val, y_val, X_test) {
  model = lm(aa_data$y ~ ., data = data.frame(aa_data$X))
  
  X_df = data.frame(X_train)
  res_tr = sapply(1:nrow(X_train), function (row_i) { 
    predict(model, X_df[row_i, ])
  })
  
  X_df = data.frame(X_val)
  res_val = sapply(1:nrow(X_val), function (row_i) { 
    predict(model, X_df[row_i, ])
  })
  
  merged = rbind(X_train, X_val)
  merged_y = c(y_train, y_val)
  X_df = data.frame(merged)
  model = lm(merged_y ~ ., data = X_df)
  imp_tst = sapply(1:nrow(X_test), function (row_i) { 
    predict(model, X_df[row_i, ])
  })
  
  list(res_tr, res_val, imp_tst)
}


eval_model <- function (.data, .fun, ...) {
  .scorer <- function (ytrue, ypred) {
    sqrt(mean((ytrue - ypred) ^ 2))
  }
  
  res_tr = c()
  res_val = c()
  for (i in 1:length(.data$train)) {
    tmp = .fun(.data$X[.data$train[[i]], ], 
               .data$y[.data$train[[i]]], 
               .data$X[.data$val[[i]], ], 
               .data$y[.data$val[[i]]], 
               .data$X_test, ...)
    res_tr  = c(res_tr,  .scorer(tmp[[1]], .data$y[.data$train[[i]]]))
    res_val = c(res_val, .scorer(tmp[[2]], .data$y[.data$val[[i]]]))
  }
  list(tr = res_tr, val = res_val)
}
```

Model comparison **(?)**

```{r}
imp_res = list()
imp_res[["mean_1"]] = eval_model(aa_data, impute_knn_mean, .k = 1)
imp_res[["mean_3"]] = eval_model(aa_data, impute_knn_mean, .k = 3)
imp_res[["mean_5"]] = eval_model(aa_data, impute_knn_mean, .k = 5)
imp_res[["dist_1"]] = eval_model(aa_data, impute_knn_dist, .k = 1)
imp_res[["dist_3"]] = eval_model(aa_data, impute_knn_dist, .k = 3)
imp_res[["dist_5"]] = eval_model(aa_data, impute_knn_dist, .k = 5)
imp_res[["reg"]] = eval_model(aa_data, impute_reg)
imp_res = melt(imp_res)
colnames(imp_res) = c("val", "data", "app")
qplot(x = data, y = val, fill = app, data = imp_res, geom = "boxplot") + theme_bw()
```

Impute values and show new contact probability coefficient matrix

```{r}
melted = melt(aa_pair_mat)[melt(upper.tri(aa_pair_mat, T))[,3],]
melted[,1] = as.character(melted[,1])
melted[,2] = as.character(melted[,2])
test_data = melted[is.na(melted[,3]), ]

imputed = impute_reg(aa_data$X[aa_data$train[[1]], ],
                          aa_data$y[aa_data$train[[1]]],
                          aa_data$X[aa_data$val[[1]], ],
                          aa_data$y[aa_data$val[[1]]],
                          aa_data$X_test)[[3]]

aa_pair_mat_imp = aa_pair_mat
aa_pair_vec = 
for (r in 1:nrow(test_data)) {
  aa_pair_mat_imp[test_data[r,1], test_data[r,2]] = imputed[r]
  aa_pair_mat_imp[test_data[r,2], test_data[r,1]] = imputed[r]
  
  ind = intersect(which(df.aa.coef[,2] == test_data[r,1]), which(df.aa.coef[,3] == test_data[r,2]))
  if (length(ind) == 0) {
    ind = intersect(which(df.aa.coef[,2] == test_data[r,2]), which(df.aa.coef[,3] == test_data[r,1]))
  }
  df.aa.coef[ind, 1] = imputed[r]
}

# update the df.aa.coef.diag
df.aa.coef.diag = df.aa.coef
df.aa.coef.diag$aa_pair = with(df.aa.coef.diag, 
  as.factor(ifelse(as.character(aa_tcr) < as.character(aa_antigen), paste(aa_tcr, aa_antigen, sep = "_"), paste(aa_antigen, aa_tcr, sep = "_")))) 
df.aa.coef.diag = df.aa.coef.diag %>% select(aa_pair, coef)
df.aa.coef.diag = df.aa.coef.diag[!duplicated(df.aa.coef.diag[,1]),]

p2 = heatmap.2(aa_pair_mat_imp, 
          hclustfun = function(x) hclust(x, method = "ward.D2"),
          RowSideColors = aa_colors[rownames(aa_pair_mat)], 
          ColSideColors = aa_colors[colnames(aa_pair_mat)],
          trace = "none",
          #breaks = seq(-16, -7, length.out = 101),
          col=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32))
```

## Contact energies

Compute mean GROMACS energies

```{r}
df.energies = df[contact == T, .(energy.mean = mean(ifelse(energy > 0 , 0, energy))), by = "aa_pair"]
df.energies$aa_tcr = str_split_fixed(as.character(df.energies$aa_pair), "_", 2)[, 1]
df.energies$aa_antigen = str_split_fixed(as.character(df.energies$aa_pair), "_", 2)[, 2]

df.energies.tmp = df.energies
df.energies.tmp$aa_tcr = df.energies$aa_antigen
df.energies.tmp$aa_antigen = df.energies$aa_tcr

df.energies = rbind(df.energies, df.energies.tmp) %>% unique()

# transform to matrix and plot heatmap.2

aa_pair_energy_mat = dcast(df.energies, aa_tcr ~ aa_antigen, value.var = "energy.mean", fun.aggregate = mean)
rownames(aa_pair_energy_mat) = aa_pair_energy_mat$aa_tcr
aa_pair_energy_mat$aa_tcr = NULL
aa_pair_energy_mat = as.matrix(aa_pair_energy_mat)

heatmap.2(aa_pair_energy_mat, 
          hclustfun = function(x) hclust(x, method = "ward.D2"),
          RowSideColors = aa_colors[rownames(aa_pair_mat)], 
          ColSideColors = aa_colors[colnames(aa_pair_mat)],
          trace = "none",
          breaks = seq(-40, 2, length.out = 33),
          col=rev(colorRampPalette(brewer.pal(9, 'YlOrRd'))(32)))
```

### Impute contact energies

Same as above, model with Kidera factors and show

```{r, fig.width=10}
aa_data = generate_data(aa_pair_energy_mat, .cv = 10)

suppressWarnings( {
  ps = lapply(1:10, function (i) { qplot(aa_data$X[,i], aa_data$y, geom = c("point", "smooth"), method="lm") + theme_bw() })
  })
summary(lm(aa_data$y ~ aa_data$X))
do.call(gridExtra::grid.arrange, c(ps, list(ncol = 5)))
```

Model comparison **(?)**

```{r}
imp_res = list()
imp_res[["mean_1"]] = eval_model(aa_data, impute_knn_mean, .k = 1)
imp_res[["mean_3"]] = eval_model(aa_data, impute_knn_mean, .k = 3)
imp_res[["mean_5"]] = eval_model(aa_data, impute_knn_mean, .k = 5)
imp_res[["dist_1"]] = eval_model(aa_data, impute_knn_dist, .k = 1)
imp_res[["dist_3"]] = eval_model(aa_data, impute_knn_dist, .k = 3)
imp_res[["dist_5"]] = eval_model(aa_data, impute_knn_dist, .k = 5)
imp_res[["reg"]] = eval_model(aa_data, impute_reg)
imp_res = melt(imp_res)
colnames(imp_res) = c("val", "data", "app")
qplot(x = data, y = val, fill = app, data = imp_res, geom = "boxplot") + theme_bw()
```

Impute values and plot new contact energy matrix

```{r}
melted = melt(aa_pair_energy_mat)[melt(upper.tri(aa_pair_energy_mat, T))[,3],]
melted[,1] = as.character(melted[,1])
melted[,2] = as.character(melted[,2])
test_data = melted[is.na(melted[,3]), ]

imputed = impute_reg(aa_data$X[aa_data$train[[1]], ],
                          aa_data$y[aa_data$train[[1]]],
                          aa_data$X[aa_data$val[[1]], ],
                          aa_data$y[aa_data$val[[1]]],
                          aa_data$X_test)[[3]]

test_data[,3] = imputed

aa_pair_energy_mat_imp = aa_pair_energy_mat
for (r in 1:nrow(test_data)) {
  aa_pair_energy_mat_imp[test_data[r,1], test_data[r,2]] = test_data[r,3]
  aa_pair_energy_mat_imp[test_data[r,2], test_data[r,1]] = test_data[r,3]
  ind = intersect(which(df.energies[,3] == test_data[r,1]), which(df.energies[,4] == test_data[r,2]))
  if (length(ind) == 0) {
    ind = intersect(which(df.energies[,3] == test_data[r,2]), which(df.energies[,4] == test_data[r,1]))
  }
  if (length(ind) == 0) {
    df.energies = rbind(df.energies, list(paste0(test_data[r,1], "_", test_data[r,2]), test_data[r,3], test_data[r,1], test_data[r,2]))
  } else {
    df.energies[[ind, 2]] = imputed[r]
  }
}

heatmap.2(aa_pair_energy_mat_imp, 
          hclustfun = function(x) hclust(x, method = "ward.D2"),
          RowSideColors = aa_colors[rownames(aa_pair_mat)], 
          ColSideColors = aa_colors[colnames(aa_pair_mat)],
          trace = "none",
          breaks = seq(-40, 2, length.out = 33),
          col=rev(colorRampPalette(brewer.pal(9, 'YlOrRd'))(32)))
```

# Building and testing a predictor

Append amino acid distance coefficients

```{r}
df.pred = df.pred[df.aa.coef.diag, on = "aa_pair"]
df.pred.trimmed = df.pred[distance_CA <= 15] # discard AAs that are too far away for training
```

Final GLM to fit contacts, we use it to combine our coefficients (computed as ``contact ~ distance_CA + coef``) with mean distances (``distance_CA.m``, some approximation of ``distance_CA``):

```{r}
# Train on a trimmed dataset
contact_glm = glm(contact ~ distance_CA.m + coef, family = binomial(), data = df.pred.trimmed)

summary(contact_glm)

print(contact_glm$coefficients)

df.pred.trimmed$p = predict(contact_glm, df.pred.trimmed, type="response")
df.pred$p = predict(contact_glm, df.pred, type="response")

df.pred = df.pred[df.energies, on = .(aa_tcr, aa_antigen)]

#pp = -with(df.pred, contact_glm$coefficients["(Intercept)"] + 
#  contact_glm$coefficients["distance_CA.m"] * distance_CA.m + 
# contact_glm$coefficients["coef"] * coef)
#pp = 1/(1+exp(pp))
#z = df.pred$p - pp
```

Save model for further evaluation:

```{r}
save(df.ca.mean, aa_pair_mat_imp, aa_pair_energy_mat_imp, contact_glm, file="../evaluation/pcm_local_glm.RData")
#write.table(df.ca.mean, "../evaluation/ca_dist_mean.txt", sep="\t", quote=F, row.names = F)
#write.table(df.aa.coef, "../evaluation/aa_pairwise_contact_coef.txt", sep="\t", quote=F, row.names = F)
#write.table(df.energies, "../evaluation/aa_pairwise_energy.txt", sep="\t", quote=F, row.names = F)
```

## Check accuracy

### General

ROC curve

```{r}
rocobj = plot.roc(as.data.frame(df.pred.trimmed)[,"contact"], df.pred.trimmed$p, ci=T)

rocobj
```

Compute true and estimated total number of contacts

```{r}
df.pred.contsum = df.pred.trimmed[, .(total_contacts = sum(contact), pred_contacts = sum(p, na.rm=T)), by = .(pdb_id, tcr_region, mhc_type, tcr_chain, len_tcr)]

ggplot(df.pred.contsum, aes(x=pred_contacts, y=total_contacts, color = tcr_region)) + 
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  
  scale_x_continuous(limits=c(0,15)) +
  scale_y_continuous(limits=c(0,15)) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~len_tcr) +
  theme_bw()

lfit = lm(total_contacts ~ pred_contacts + len_tcr + tcr_region - 1, df.pred.contsum)
summary(lfit)
anova(lfit)
```

### Examples from the training set

Check for a couple of antigens, GLCTLVAML

```{r}
df.pred.glc = df.pred %>%
  filter(antigen_seq == "GLCTLVAML") %>%
  droplevels()

ggplot(df.pred.glc, aes(x=pos_tcr, y=pos_antigen)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p), cex=1.3) +
  geom_point(aes(color=contact)) +
  scale_x_continuous(breaks=0:20) +
  scale_y_continuous(breaks=0:20) +
  #scale_fill_gradient("P", 
  #                    low="white", high="#045a8d") +
  scale_color_manual(values = c(NA, "red")) +
  scale_fill_gradientn(colors=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32)) +
  facet_grid(tcr_chain ~ tcr_region, scales="free", space="free") +
  theme_bw()
```

and LLFGYPVAV

```{r}
df.pred.llf = df.pred %>%
  filter(antigen_seq == "LLFGYPVAV") %>%
  droplevels()

ggplot(df.pred.llf, aes(x=pos_tcr, y=pos_antigen)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p), cex=1.3) +
  geom_point(aes(color=contact)) +
  scale_x_continuous(breaks=0:20) +
  scale_y_continuous(breaks=0:20) +
  #scale_fill_gradient("P", 
  #                    low="white", high="#045a8d") +
  scale_color_manual(values = c(NA, "red")) +
  scale_fill_gradientn(colors=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32)) +
  facet_grid(tcr_chain ~ tcr_region, scales="free", space="free") +
  theme_bw()
```

# Independent validation

Computing contact map from fitted model for a specified TCR:pMHC setup

```{r}
compute_contact_map = function(mhc_type, tcr_chain, tcr_region, cdr_seq, ag_seq, id = "tmp") {
  cdr_seq = as.character(cdr_seq)
  ag_seq = as.character(ag_seq)
  df.cdr = data.frame(aa_tcr = strsplit(cdr_seq, "")[[1]], 
                      pos_tcr = 1:nchar(cdr_seq) - 1)
  df.ag = data.frame(aa_antigen = strsplit(ag_seq, "")[[1]], 
                      pos_antigen = 1:nchar(ag_seq) - 1)
  
  df.pairs = expand.grid(df.cdr$pos_tcr, df.ag$pos_antigen)
  colnames(df.pairs) = c("pos_tcr", "pos_antigen")
  
  df.pairs = merge(df.pairs, df.cdr)
  df.pairs = merge(df.pairs, df.ag)
  
  df.pairs$aa_pair = with(df.pairs, 
                       as.factor(ifelse(as.character(aa_tcr) < as.character(aa_antigen),
                                        paste(aa_tcr, aa_antigen, sep = "_"), paste(aa_antigen, aa_tcr, sep = "_")))) 
  
  df.pairs$mhc_type = mhc_type
  df.pairs$tcr_chain = tcr_chain
  df.pairs$tcr_region = tcr_region
  df.pairs$len_tcr = nchar(cdr_seq)
  df.pairs$len_antigen = nchar(ag_seq)
  
  df.pairs$pos_tcr_c = with(df.pairs, pos_tcr - round(len_tcr / 2))
  df.pairs$pos_antigen_c = with(df.pairs, pos_antigen - round(len_antigen / 2))
  
  df.pairs$id = id # ! id can be anything to group the complex, e.g. clonotype id in sample
    
  df.res = merge(df.pairs %>% select(id, mhc_type, tcr_chain, tcr_region, pos_tcr_c, pos_antigen_c, aa_antigen, aa_tcr, aa_pair), 
        df.pred %>% select(mhc_type, tcr_chain, tcr_region, pos_tcr_c, pos_antigen_c, aa_pair, p, energy.mean), all.x = T) %>%
        unique()
  
  df.res$p[is.na(df.res$p)] = 0
  df.res$energy.mean[is.na(df.res$energy.mean)] = 0
  
  df.res
}
```

## Example 1

Testing on data from [http://www.jbc.org/content/289/2/628.full](REF1) - engeneered peptide

```{r}
cm.tmp = compute_contact_map("MHCI","TRA","CDR3","CAVDSATSGTYKYIF","ILAKFLHWL","WT")
cm.tmp = rbind(cm.tmp, compute_contact_map("MHCI","TRA","CDR3","CAVDSATSGTYKYIF","ILAAFLHWL","No binding"))
cm.tmp = rbind(cm.tmp, compute_contact_map("MHCI","TRA","CDR3","CAVDSATSGTYKYIF","GLGGGGGGV","Mock"))
cm.tmp = rbind(cm.tmp, compute_contact_map("MHCI","TRA","CDR3","CAVDSATALPYGYIF","ILAKFLHWL", "Enhanced"))

print(cm.tmp %>% group_by(id) %>% summarize(contacts = sum(p), energy = sum(p * energy.mean)))

ggplot(cm.tmp, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("P", colors=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32)) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()

ggplot(cm.tmp, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p * energy.mean), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("E", colors=rev(colorRampPalette(brewer.pal(9, 'YlOrRd'))(32))) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()
```

## Example 2

Testing on data from [http://www.nature.com/articles/ncomms6223](REF2) - high affinity Tax mutant

```{r}
cm.tmp = compute_contact_map("MHCI","TRB","CDR3","CASRPGLAGGRPEQYF","LLFGYPVYV","WT")
cm.tmp = rbind(cm.tmp, compute_contact_map("MHCI","TRB","CDR3","CASRPGLMSAQPEQLF","LLFGYPVYV","High affinity (A6-X15)"))
cm.tmp = rbind(cm.tmp, compute_contact_map("MHCI","TRB","CDR3","CASRPGWMSAQPEQLF","LLFGYPVYV","Attenuated A6-X15 1"))
cm.tmp = rbind(cm.tmp, compute_contact_map("MHCI","TRB","CDR3","CASRPGMMSAQPEQLF","LLFGYPVYV","Attenuated A6-X15 2"))

print(cm.tmp %>% group_by(id) %>% summarize(contacts = sum(p), energy = sum(p * energy.mean)))

ggplot(cm.tmp, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("P", colors=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32)) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()

ggplot(cm.tmp, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p * energy.mean), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("E", colors=rev(colorRampPalette(brewer.pal(9, 'YlOrRd'))(32))) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()
```

## Example 3

Continue with REF2: comparing Tax specific vs Tax and MART specific vs MART + cross-comparison. We compare CDR3beta of A6 (wild-type Tax-specific variant) and a MART-specific TCR derived from A6 by direct evolution.

```{r}
cm.tmp.1 = compute_contact_map("MHCI","TRB","CDR3","CASRPGLAGGRPEQYF","LLFGYPVYV","Tax-tcr:Tax")
cm.tmp.1 = rbind(cm.tmp.1, compute_contact_map("MHCI","TRB","CDR3","CASRPGLAGGRPEQYF","AAGIGILTV","Tax-tcr:MART"))
cm.tmp.1 = rbind(cm.tmp.1, compute_contact_map("MHCI","TRB","CDR3","CASRPGWMAGGVELYF","LLFGYPVYV","MART-tcr:Tax"))
cm.tmp.1 = rbind(cm.tmp.1, compute_contact_map("MHCI","TRB","CDR3","CASRPGWMAGGVELYF","AAGIGILTV","MART-tcr:MART"))

print(cm.tmp.1 %>% group_by(id) %>% summarize(contacts = sum(p), energy = sum(p * energy.mean)))

ggplot(cm.tmp.1, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("P", colors=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32)) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()

ggplot(cm.tmp.1, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p * energy.mean), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("E", colors=rev(colorRampPalette(brewer.pal(9, 'YlOrRd'))(32))) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()
```

Same as above for alpha chains

```{r}
cm.tmp.2 = compute_contact_map("MHCI","TRA","CDR3","CAVVVTTDSWGKLQF","LLFGYPVYV","Tax-tcr:Tax")
cm.tmp.2 = rbind(cm.tmp.2, compute_contact_map("MHCI","TRA","CDR3","CAVVVTTDSWGKLQF","AAGIGILTV","Tax-tcr:MART"))
cm.tmp.2 = rbind(cm.tmp.2, compute_contact_map("MHCI","TRA","CDR3","CAVVVTKYSWGKLQF","LLFGYPVYV","MART-tcr:Tax"))
cm.tmp.2 = rbind(cm.tmp.2, compute_contact_map("MHCI","TRA","CDR3","CAVVVTKYSWGKLQF","AAGIGILTV","MART-tcr:MART"))

print(cm.tmp.2 %>% group_by(id) %>% summarize(contacts = sum(p), energy = sum(p * energy.mean)))

ggplot(cm.tmp.2, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("P", colors=colorRampPalette(brewer.pal(9, 'YlGnBu'))(32)) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()

ggplot(cm.tmp.2, aes(x=pos_tcr_c, y=pos_antigen_c)) +
  geom_tile(fill=NA) +
  geom_label(aes(label=paste(aa_tcr, aa_antigen, sep=":"), fill = p * energy.mean), cex=2) +
  scale_x_continuous(breaks=-8:9) +
  scale_y_continuous(breaks=-5:5) +
  scale_fill_gradientn("E", colors=rev(colorRampPalette(brewer.pal(9, 'YlOrRd'))(32))) +
  facet_grid(id~., scales="free", space="free") +
  theme_bw()
```

Summarize contacts and energies for TRA and TRB:

```{r}
print(rbind(cm.tmp.1, cm.tmp.2) %>% group_by(id) %>% summarize(contacts = sum(p), energy = sum(p * energy.mean)))
```

## Example 3

> Pehaps can also add https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2681418/


